# Creating a dataset

Code to shrink the dataset with all of the wanding and environmental data into a more manageable dataset with just the Obear brook observations and relevant explanatory variables.

## Running Code

Load the dataset with all of the wanding and environmental data into R.
```{r load in data table}
cdWB <- read.csv('/Users/jpilchik/OneDrive - DOI/Documents/Wanding_Data/data/cdWB_All.csv', header=T)
```

Filter the dataset to include only the Obear brook observations.
```{r filter to just Obear}
library(dplyr)

#specify team names to keep
river_name <- 'wb obear'

#select all rows where team is in list of team names to keep
obear_All <- cdWB %>% filter(river %in% river_name)
head(obear_All)
```

Break the date observations into month and year so that those variables can be used separately as explanatory variables.
```{r}
library(tidyverse)
obear <- separate(obear_All, detectionDate, into = c('year', 'month', 'day', 'time')) %>% dplyr::select(-one_of('day','time'))
head(obear)
```

Filter the dataset to include only the relevant explanatory variables.
```{r}
obear <- obear %>% dplyr::select(tag, year, month, river, section, observedLength, observedWeight, meanTemperature, meanFlowByRiver, isYOY, distMoved, moveDir)
head(obear)
```
```{r}
# Remove NA values
sub <- obear[!is.na(obear$distMoved),]
sub <- sub[!is.na(sub$observedLength),]
sub <- sub[!is.na(sub$observedWeight),]
head(sub)
```

```{r}
# Separate adults from YOY
sub$isYOY <- as.factor(sub$isYOY)
adult_all <- sub[ which(sub$isYOY=='FALSE'), ]
```

```{r}
# Remove the -90 value
adult <- adult_all[-which(adult_all$distMoved==-90), ]

# Create a new distMovedp column that removes negative values
adult$distMovedp <- abs(adult$distMoved)

#rescale explanatory variables
adult$observedWeightz <- c(scale(adult$observedWeight))
adult$meanTemperaturez <- c(scale(adult$meanTemperature))
adult$meanFlowByRiverz <- c(scale(adult$meanFlowByRiver))
adult$observedLengthz <- c(scale(adult$observedLength))

# Turn categorical variables into factors
adult$year <- as.factor(adult$year)
adult$month <- as.factor(adult$month)
adult$section <- as.factor(adult$section)

adult$tag <- trimws(adult$tag)
# Convert 'tag' to a factor to assign unique levels
adult$tag_factor <- factor(adult$tag)
# Convert the factor levels to sequential ID numbers
adult$tagID <- as.numeric(adult$tag_factor)
adult$tagID_factor <- factor(adult$tagID)

head(adult)
summary(adult)
```

